---
title: "Estimate features for Dis3L2 homologs"
author: "Edward Wallace"
date: "17/07/2020"
output: html_document
---

# Summary

This file calculates information and features of Dis3L2/Ssd1 homologs, from multiple sequence alignments in `data/Dis3L2_only`, files `Dis3L2_BLASTp_select_fullseqs_mafftlocal.fasta`.
These features include:

* sequence ids
* species and summary taxonomic information
* presence of conserved RNaseII active site
* Cell-wall biogenesis kinase Cbk1 consensus phosphorylation site and docking site
* Cyclin-dependent kinase Cdk1 consensus phosphorylation site 
* predicted nuclear localisation sequence


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here)
library(tidyverse)
library(Biostrings)
```

## Load Dis3L2 sequences

```{r load_seqs}
Dis3L2_Ntermaln <- 
    here::here(
        "data",
        "Dis3L2_only",
        "Dis3L2_BLASTp_select_fullseqs_mafftlocal_Nterm1to337.fasta") %>%
    Biostrings::readAAStringSet()

Dis3L2_Ntermseqs <- Dis3L2_Ntermaln %>%
    as.character %>%
    stringr::str_remove_all("-")

Dis3L2_fullaln <-
    here::here(
        "data",
        "Dis3L2_only",
        "Dis3L2_BLASTp_select_fullseqs_mafftlocal.fasta") %>%
    Biostrings::readAAStringSet()

Dis3L2_fullseqs <- Dis3L2_fullaln %>% 
    as.character %>%
    stringr::str_remove_all("-")

stopifnot(all.equal(names(Dis3L2_Ntermaln),names(Dis3L2_fullaln)))

Dis3L2_trimal_alnseqs <- 
    here::here(
        "data",
        "Dis3L2_only",
        "Dis3L2_BLASTp_select_fullseqs_mafftlocal_trimal.fasta") %>%
    Biostrings::readAAStringSet()
```

## Find Cbk1 phosphorylation sites in N-terminus

Search for `H[RK]xxx[ST]` or `Hx[RK]xx[ST]` motifs of Cbk1 cell wall biogenesis kinase.

```{r find_cbk1_nterm,dependson="load_seqs"}
cbk1cons_alt <- "H[A-Z]([A-Z][RK]|[RK][A-Z])[A-Z][ST]"
cbk1dock <- "[YF][A-Z]FP"
cbk1cons_df <-
    tibble(ID = names(Dis3L2_Ntermaln) %>% 
               stringr::str_extract("[\\w.]+"),
        cbk1cons_count = Dis3L2_Ntermseqs %>% 
            stringr::str_count(cbk1cons_alt),
        cbk1dock_count = Dis3L2_Ntermseqs %>% 
            stringr::str_count(cbk1dock))

cbk1cons_df
```

## Find Cdk1 phosphorylation sites in N-terminus

Search for `[ST]Px[KR]` motif of Cdk1 cyclin-dependent kinase.

```{r find_cdk1,dependson="load_seqs"}
cdk1cons <- "[ST]P[A-Z][KR]"
cdk1consdock <- "[ST][QP][A-Z]{5,20}[KR][A-Z]L"
cdk1cons_df <-
    tibble(ID = names(Dis3L2_fullaln) %>% 
               stringr::str_extract("[\\w.]+"),
        cdk1cons_count = Dis3L2_fullseqs %>% 
            stringr::str_count(cdk1cons),
        cdk1consdock_count = Dis3L2_fullseqs %>% 
            stringr::str_count(cdk1consdock),
        cdk1cons_Ntcount = Dis3L2_Ntermseqs %>% 
            stringr::str_count(cdk1cons),
        cdk1consdock_Ntcount = Dis3L2_Ntermseqs %>% 
            stringr::str_count(cdk1consdock))
cdk1cons_df
```

## Score for active site

Score presence of the active site `D[A-Z]{7}DD`, within the 20 nts up/down of the visually-inspected S. pombe Dis3L2 active site in the trimmed alignment.

```{r score_activesite,dependson="load_seqs"}
activesite = "D[A-Z]{7}DD"
activesite_pos <- 1174

activesite_df <-
    tibble(ID = names(Dis3L2_trimal_alnseqs) %>% 
               stringr::str_extract("[\\w.]+"),
        activesite = as.character(Dis3L2_trimal_alnseqs) %>% 
            stringr::str_sub(start = activesite_pos-20, end=activesite_pos+20) %>%
            stringr::str_detect(activesite))

activesite_df
```

## Search for nuclear localisation sites 

Generic monopartite/bipartite nuclear localisation sites from Kosugi et al., doi: 10.1074/jbc.M807017200

Ssd1-specific nuclear localisation sites developed here by inspection of multiple sequence alignment, following Kurischko et al, doi: 10.1111/j.1365-2958.2011.07731.x.

Searching in trimmed alignment

```{r search_NLS,dependson="load_seqs"}
NLS_regex_bipartite <- "[KR]{2}[A-Z]{9,12}[KR]{3,5}"
NLS_regex_monopartite <- "[KR]{4,6}"
NLS_regex_Ssd1like <- "([KRE]{2,4}[ADHILMNVSPQ]{0,3}){2,5}"
NLS_pos_Ssd1like <- 894

NLS_df <-
    tibble(ID = names(Dis3L2_trimal_alnseqs) %>% 
               stringr::str_extract("[\\w.]+"),
        NLS_bi_count = as.character(Dis3L2_trimal_alnseqs) %>% 
            stringr::str_remove_all("-") %>%
            stringr::str_count(NLS_regex_bipartite),
        NLS_mono_count = as.character(Dis3L2_trimal_alnseqs) %>% 
            stringr::str_remove_all("-") %>%
            stringr::str_count(NLS_regex_monopartite),
        NLS_Ssd1like = as.character(Dis3L2_trimal_alnseqs) %>% 
            stringr::str_sub(start = NLS_pos_Ssd1like-20, end=NLS_pos_Ssd1like+20) %>%
            stringr::str_remove_all("-") %>%
            stringr::str_detect(NLS_regex_Ssd1like))
NLS_df
```

## Collect into features data frame

```{r ids_dataframe,dependson="load_seqs"}
Dis3L2_IDs_df <- 
    tibble(ID = names(Dis3L2_fullaln) %>% 
               stringr::str_extract("[\\w.]+"),
           Organism = names(Dis3L2_fullaln) %>% 
               stringr::str_extract("\\[.*\\]") %>%
               stringr::str_extract("\\w+ \\w+"),
           Info = names(Dis3L2_fullaln) )
```

## Nicely describe taxonomic info

```{r add_taxon_info}
add_extensions <- function(x) {
    if( all(is.na(x)) & length(x) > 1) {
        x <- as.character(1:length(x))
    }
    return(x)
}

Dis3L2_clade_df <- 
    here::here(
        "data",
        "Dis3L2_only",
        "Dis3L2_BLASTp_select_taxa.txt") %>%
    read_tsv(comment="#") %>%
    select(ID,Organism,Clade,Extension) %>%
    group_by(Organism) %>%
    mutate(Extension = add_extensions(Extension))
```

## Combine information into a single table and write to a file.

```{r combine_data,dependson=c("ids_dataframe","add_taxon_info","score_activesite","find_cbk1_nterm","search_NLS")}
Dis3L2_features_file <- here::here("data",
                                   "Dis3L2_only",
                                   "Dis3L2_BLASTp_select_features.txt")

Dis3L2_features_df <- Dis3L2_IDs_df %>%
    left_join(Dis3L2_clade_df) %>%
    left_join(activesite_df) %>%
    left_join(cbk1cons_df) %>%
    left_join(cdk1cons_df) %>%
    left_join(NLS_df)

write_lines(c("# Dis3L2_BLASTp_select_features.txt",
            "# Information and features of Dis3L2/Ssd1 homologs ",
            "# calculated from Dis3L2_BLASTp_select_fullseqs_mafftlocal.fasta, etc.",
            "# taxonomic info, active site, cbk1 consensus phosphorylation site, cdk1 sites, nuclear localisation sequence",
            "# see estimate_features_Dis3L2_only.Rmd for full descriptions and code",
            "# Edward.Wallace@ed.ac.uk, 17 July 2020",
            "# "),
            path=Dis3L2_features_file)
write_tsv(Dis3L2_features_df, path=Dis3L2_features_file, 
          col_names = TRUE, append=TRUE)
```
